---
title: "Dataviz with ggplot2"
author: "Marcelino Guerra"
date: "Last update: 2/4/2021"
abstract: Lecture notes on how to make line plots and scatter plots with ggplot2, and an exercise with tidycensus.  
output: 
  rmdformats::downcute
---

<style type="text/css">
  body{
  font-size: 13pt;
}
</style>

```{r setup, include=FALSE}
  knitr::opts_chunk$set(include = TRUE)  # TRUE for solution; FALSE for questions set
  knitr::opts_chunk$set(echo = TRUE)
  knitr::opts_chunk$set(message = FALSE)
  knitr::opts_chunk$set(warning = FALSE)
  knitr::opts_chunk$set(fig.height = 6, fig.width = 11, out.width = '100%', fig.align = "center")
  options(width = 90)

```


## Line plots

In this exercise, we are working with urbanization and per capita income of a bunch of countries from [Our World in Data](https://ourworldindata.org/urbanization). The first chart is a simple line plot. The first step is to import [this .csv file](https://drive.google.com/file/d/1M13_5sjCNnouvGZOJB56SemHydPkgCvR/view?usp=sharing). You might want to check the data using `View()`.

```{r,include=T}
library(tidyverse)
urban_gdp<-read_csv("urbanization_gdp.csv", col_names=T)
```

Let's check the evolution of the share of urban population (%) in Brazil, from 1950 to 2016. To do that, you need to filter the full dataset.

```{r, include=T}
brazil<-urban_gdp%>%filter(Entity=="Brazil")%>%arrange(Year)%>%filter(Year>=1950)
```

Working with the filtered data, use `ggplot()` package to create line plots. The package `ggthemes` allows you to customize the chart appearance, and you might want to check all the possibilites [here](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/). For this exercise, I will use [`theme_economist`](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/theme_economist/). The basic line plot works with `geom_line()`, and you can set the line type and its size easily. `theme()` adjust the size of the label in the axis, and `scale_x_continuous()` gives you some freedom to establish the years that appear in the x axis (note that we are not working with dates yet!).  

```{r, include=T}
library(ggthemes)

ggplot(data=brazil, aes(x=Year,y=`Share of population living in urban areas (%)`))+
  geom_line(size=1.4, color="skyblue4")+ labs(x = "Year", y="Share of population living in urban areas (%) in Brazil")+
  theme_economist(base_size = 14) +scale_colour_economist()+
  scale_x_continuous(breaks = seq(from = 1950, to =2016 , by =10))+
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=12,face="bold"))
```

## Scatter plots

Using a scatter plot, you can see the extent to which two variables are associated, and it is a very compelling way to show simple correlations. For example, one can argue that *per capita* income goes along with urbanization - as countries urbanize, they get richer. To construct a scatter plot between those two variables for 2016, we filter the full dataset (`urban_gdp`) for `Year==2016`. You can check your new dataset using `View(urban_gdp)`. There are entities without information about `GDP per capita`, and we drop those NA's. If you look at the data, you will realize that when an Entity has `Code` equal to NA that means the entity is not actually a single country, and we also want to drop those data points. Finally, we drop the entity `World` because we only want to work with countries.     

As you can see, cleaning the data is a crucial step before plotting. Finally, I will rename columns 4 and 5 and create a new variable applying the natural logarithm to the GDP per capita. 

```{r, include=T}
urban_gdp_16<-urban_gdp%>%filter(Year==2016)
urban_gdp_16<-urban_gdp_16%>%drop_na(Code, `GDP per capita (2011 int-$) ($)`)
urban_gdp_16<-urban_gdp_16%>%filter(Entity!="World")
names(urban_gdp_16)[4]<-"urban_pop"
names(urban_gdp_16)[5]<-"gdp"
urban_gdp_16$ln_gdp<-log(urban_gdp_16$gdp)
head(urban_gdp_16,3)
```

Back to `ggplot()`! Instead of `geom_line()`, we use `geom_point()` to construct scatter plots. 

```{r, include=T}
ggplot(urban_gdp_16, aes(x=urban_pop, y=ln_gdp)) +
  geom_point(color="darkgrey") + 
  scale_x_continuous(name = "Share of population living in urban areas (%)") +
  scale_y_continuous(name = "Ln GDP per capita U$ 2011") +
  theme_economist_white(base_size = 17, gray_bg=FALSE)+
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=12,face="bold"))
```

**You can do better than that**. The package `ggrepel` allows you to label the dots in your chart (function `geom_text_repel()`). I will label the dots with respective country codes. Finally, `stat_smooth()` adds the regression line to your plot.

```{r}
library(ggrepel)
ggplot(urban_gdp_16, aes(x=urban_pop, y=ln_gdp)) + geom_point(color="darkgrey") + geom_text_repel(aes(label=Code), size=3)+
  stat_smooth(method = "lm", formula =y~x, se=F, color="black") +
  scale_x_continuous(name = "Share of population living in urban areas (%)") +
  scale_y_continuous(name = "Ln GDP per capita U$ 2011") +
  theme_economist_white(base_size = 17, gray_bg=FALSE)+
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=12,face="bold"))
```

Here we see the strong and positive relationship between the share of the population living in urban areas (%) and the GDP per capita (U$ 2011) for 164 selected countries. **For more facts about urbanization, check [Our World in Data](https://ourworldindata.org/urbanization)**.